A Switching Planner for Combined Task and Observation Planning

### 2010-11-21 subarchitectures/planner.sa/trunk/src/latex/ICAPS-2011/moritz_2011.tex
Paper do far
Initial state description to actions, sketched.
 - Charels

Any (PO)MDP with stochastic actions can be compiled into an equivalent
deterministic-action POMDP with all the action uncertainty expressed
in the starting-state distribution~\cite{ng:Jordan:2000}. In our
setting, that of finite-horizon planning, this compilation yields a
finite flat representation of the POMDP. Therefore, although our
current implementation does not explicitly support it, our approach
generalises to POMDPs with stochastic action models.

@@

A general formalism for representing solutions to POMDPs is the
finite-state controller (FSC). This is a three-tuple $\langle \nodes,
\selection, \transition \rangle$ where: $\node \in \nodes$ is a set of
nodes, $\selection_\node(\action) = P(\action | \node)$, and
$\transition_\node(\action, \observation, \node') = P(\node'|\node,
\action, \observation)$. The value of state $\state$ at node $\node$
given a problem model, written $V_\node(\state)$, is:

\begin{equation}\label{eq:evaluation}
\begin{array}{lcl}
V_\node(\state) & = & \sum_{\action \in \actions}
\selection_\node(\action) \reward(\state, \action) \;\; + \vspace{1ex} \\

&& \hspace{-10ex} \beta \sum_{\action, \observation,
\state', \node'} \transition_\node(\action, \observation, \node')
\transProb(\state, \action, \state') \obsDist_\observ(\state',
\action) V_{\node'}(\state')
\end{array}
\end{equation}

\noindent The value of $\bstate$ according to a controller is
therefore:

\begin{equation} \label{eq:valueBelief}
V_{\pp{FSC}}(\bstate) = \max_{\node \in \nodes} \sum_{\state \in \states} \bstate(\state) V_\node(\state)
\end{equation}

\section{Serial Planning in Determinisations}

Goal expressions in the problem description ---for example
$(\pp{kval}~(\pp{location}~\pp{cornflakes}))$ expresses ``the robot
knows where the {\em cornflakes}--- are treated as control-knowledge
by the planner. Here, a plan is not considered valid unless in the
finial stet the goal condition is satisfied.

@@

Because PPDDL cannot model domains that feature partial observability,
we developed an extension of that language, called DTPDDL. This can
express probabilistic models of the sensing consequences of acting, to
quantitatively capture unreliability in perception. That expressive
power is achieved by incorporating perceptual analogues of fluent,
predicate, and action definitions. In detail, we have declarations of
state characterising predicate and fluent symbols according to PPDDL
syntax. In addition, we allow two other declarative blocks, for
perceptual predicates and fluents respectively. For example, suppose our
robot is tasked with exploring {\em locations} in order to identify
the whereabouts of a {\em visual-object}. We must describe state and
perceptual facts that model the {\em true}, resp. perceived, locations
of objects. In DTPDDL, these declarations appear as:

\small{
\begin{tabtt}
(\=:functions  ;; state fluents\\
  \> (is-in ?v - visual-object) - location )\\
(:observable-functions  ;; perceptual fluents\\
  \> (o-is-in ?v - visual-object) - location )
\end{tabtt}
}

\noindent To model the sensing capabilities of the agent, we have
operator-like descriptive elements, with preconditions in both action
and state symbols, and uniformly positive effects over perceptual
predicates -- I.e, the negation normal form of a sensing effect
formula cannot contain negation. Whereas the effects of an {\em
operator} schema are over state propositions, and describe how the
states change under application of actions, the effects of a {\em
sense} schema are perceptual, and describe the composition of an
observation following the execution of an action. Also, and without a
loss of generality, we find it convenient to separate the actional
precondition from the state precondition. For example, our robot is
able to look for a visual object, such as a box of cornflakes, at a
given place. We can model that deterministic action using the
following operator schema:

\small{
\begin{tabtt}
(\=:action look-for-object \+ \\
   :parameters (\=?r - robot ?v - visual-object\\
   \> ?l - location) \\
   :precondition (and (= (is-in ?r) ?l) ) \\
   :effect (and (assign (reward) -3) ) ) \\
\end{tabtt}
}

@@

The syntax and semantics for describing initial state distributions in
DTPDDL is taken verbatim from PPDDL. It is useful to present that
factored tree-like structure here, in order to aid our discussion of
sequential planning in POMDP determinisations. The distribution is
expressed in a tree-like structure of terms. Each term is either: (1)
atomic, e.g., $(=(\pp{location}~\pp{box})~\pp{office})$, (2)
probabilistic, e.g., $(\pp{probabilistic}~p_1 (T_1) .. p_n (T_n))$
where $T_i$ are conjunctive, or (3) a conjunct over probabilistic and
atomic terms. The root term is always conjunctive, and the leaves are
atomic. For example, the starting distribution for $\pp{R2D2}$ might
be described thusly:

\small{
\begin{tabtt}
(\=:init (= (is-in R2D2) kitchen) \+ \\
       (probabilistic \=.8 (= (is-in box) office)  \\
		      \>.2 (= (is-in box) kitchen)) \\
       (probabilistic .3 (= (is-in cup) office)  \\
		      \>.7 (= (is-in cup) kitchen))) \\
\end{tabtt}
}

\noindent The interpretation of such an expression can be given
according to a recursive descent. Here, an atom is visited as soon as
its conjunctive parent is visited, and a conjunctive term is visited
if all its immediate subterms are visited. A probabilistic term is
visited if exactly one of its subterms, $T_i$, is visited. Each valid
visitation of the root term according to this recursive definition
encapsulates a starting state, along with the probability it
occurs. The former corresponds to the union of all visited atoms, and
the latter corresponds to the product of $p_i$ entries on the visited
subterms of probabilistic elements. Making these ideas concrete, our
example yields the following flat state distribution:

\begin{tabular}{cccc}
\hline
Probability & (is-in R2D2)  & (is-in box)  & (is-in cup) \\
\hline
.24 & kitchen & office & office \\
.06 & kitchen & kitchen & office \\
.56 & kitchen & office & kitchen \\
.14 & kitchen & kitchen & kitchen \\
\hline
\end{tabular}

@@

\section{Discussion}

There have been a number of developments recently towards planning
under uncertainty using systems that were intended for serial planning
in deterministic problems.  Notably,
\system{FFR$_a$}~\cite{yoon:etal:2007}, the winning entry from the
probabilistic track of the 2004 International Planning Competition,
uses the fast satisficing procedure
\system{FF}~\cite{hoffmann:nebel:2001} to compute a serial plan and
corresponding execution trace. The resulting plan is then executed
until the observed trace deviates from the planned trace.

More computationally expensive approaches in this vein combine
sampling strategies on valuations over {\em runtime variables} with
deterministic planning procedures. The outcome is typically a more
robust serial plan~\cite{yoon:etal:2008}, or contingent
plan~\cite{majercik:2006}.

%%
Recently the systems
\system{Conformant-FF}~\cite{hoffmann:brafman:2006} and
$T_0$~\cite{palacios:geffner:2009}, demonstrated how conformant
planning ---i.e., serial planning in unobservable worlds--- can be
modelled as a deterministic problem, and therefore solved using serial
systems. In this conformant setting, advances have been towards
compact representations of beliefs amenable to existing best-first
search planning procedures, and lazy evaluations of beliefs. This
research thread has also been extended recently to contingent planning
in fully observable non-deterministic
environments~\cite{albore:etal:2009}.

@@

In this paper we address these challenges, developing a continual
planner that switches between sequential and decision-theoretic
planning. Given a POMDP model of the environment, serial planning is
responsible for an initial deterministic sequential plan and
complementary runtime evolution of the decision process. That plan is
executed until an assumption about a runtime proposition

@@

Continual planning is a powerful technique that goes some way to
addressing those challenges. That approach interleaves planning and
execution, deliberately postponing planing for contingencies unless
they eventuate during execution.

computing a single serial plan and
eventuality

@@

>>>>

@@

\subsection{Plan Evaluation}

An optimal solution to a finite-horizon POMDP is a contingent plan,
and can be expressed as a mapping from observation histories to
actions. Although suboptimal in general, useful plans can also take a
classical sequential format. This is the case in {\em conformant}
planning, where the objective is to find a sequence of actions that
achieves a goal ---I.e., reaches a state that satisfies some given
Boolean condition--- with probability $1$.  More generally, a useful
formalism for representing solutions to POMDPs is the finite-state
controller (FSC). This is a three-tuple $\langle \nodes, \selection,
\transition \rangle$ where: $\node \in \nodes$ is a set of controller
states, $\selection_\node(\action) = P(\action | \node)$ gives the
probability that the controller prescribes \action\ when in state
\node, and $\transition_\node(\action, \observation, \node') =
P(\node'|\node, \action, \observation)$ gives the probability the
controller transitions to state $\node'$, supposing we execute
\action\ in state \node\ receiving observation \observation. When
acting for finite $N$ steps, the value of a controller corresponds to
the expected cumulative reward:

\begin{equation}\label{eq:expectedvalue}
V_{\pp{FSC}}(\bstate) = \Expect \bigg{[} 
\sum_{t=0}^{N}  \reward(\bstate_t) \mid \pp{FSC}, \bstate = \bstate_0 \bigg{]}
\end{equation}

\noindent Where $b_t$ is the belief state at time $t$. Finally, it is
useful to note that there is a corresponding deterministic FSCs for
both sequential (e.g., conformant) and contingent plans.

@@

Because PPDDL cannot model domains that feature partial observability,
we developed an extension of that language, called DTPDDL. This can
express probabilistic models of the sensing consequences of acting, to
quantitatively capture unreliability in perception. That expressive
power is achieved by incorporating perceptual analogues of fluent,
predicate, and action definitions. In detail, we have declarations of
state characterising predicate and fluent symbols according to the
PPDDL syntax. In addition, we allow two other declarative blocks, for
perceptual predicates and fluents respectively. For example, suppose
our robot is tasked with exploring {\em locations} in order to
identify the whereabouts of a {\em visual-object}. We must describe
state and perceptual facts that model the {\em true}, resp. perceived,
locations of objects. In DTPDDL, these declarations appear as:

\small
\begin{tabtt}
(\=:functions  ;; state fluents\\
  \> (is-in ?v - visual-object) - location )\\
(:observable-functions  ;; perceptual fluents\\
  \> (o-is-in ?v - visual-object) - location )
\end{tabtt}
\normalsize

\noindent To model the sensing capabilities of the agent, we have
operator-like descriptive elements, with preconditions in both action
and state symbols, and uniformly positive effects over perceptual
predicates -- I.e, the negation normal form of a sensing effect
formula cannot contain negation. Whereas the effects of an {\em
operator} schema are over state propositions, and describe how the
states change under application of actions, the effects of a {\em
sense} schema are perceptual, and describe the composition of an
observation following the execution of an action. Also, and without a
loss of generality, we find it convenient to separate the actional
precondition from the state precondition.

Making the above ideas concrete with an example, suppose a robot is
able to look for a visual-object, such as a box, at a given place. We
can model that deterministic action using the following operator
schema:

\small
\begin{tabtt}
(\=:action look-for-object \+ \\
   :parameters (\=?r - robot ?v - visual-object\\
   \> ?l - location) \\
   :precondition (and (= (is-in ?r) ?l) ) \\
   :effect (and (assign (reward) -3) ) ) \\
\end{tabtt}
\normalsize

@@

The syntax and semantics for describing initial state distributions in
DTPDDL is taken verbatim from PPDDL. It is useful to present that
factored tree-like structure here, in order to aid us when discussing
sequential planning in POMDP determinisations. The distribution is
expressed in a tree-like structure of terms. Each term is either: (1)
atomic, e.g., $(=(\pp{location}~\pp{box})~\pp{office})$, (2)
probabilistic, e.g., $(\pp{probabilistic}~p_1 (T_1) .. p_n (T_n))$
where $T_i$ are conjunctive, or (3) a conjunct over probabilistic and
atomic terms. The root term is always conjunctive, and the leaves are
atomic. For example, the starting distribution for $\pp{R2D2}$ might
be described thusly:

\small
\begin{tabtt}
(\=:init (= (is-in R2D2) kitchen) \+ \\
       (probabilistic \=.8 (= (is-in box) office)  \\
		      \>.2 (= (is-in box) kitchen)) \\
       (probabilistic .3 (= (is-in cup) office)  \\
		      \>.7 (= (is-in cup) kitchen))) \\
\end{tabtt}
\normalsize

\noindent The interpretation of such an expression can be given
according to a recursive notion of {\em visitation} over terms. Here,
an atom is {\em visited} iff its conjunctive parent is visited, and a
conjunctive term is visited iff all its immediate subterms are
visited. A probabilistic term is visited iff its conjunctive parent is
visited, and exactly one of its subterms, $T_i$, is visited. Each
valid visitation of the root term according to this recursive
definition encapsulates a starting state, along with the probability
it occurs. The former corresponds to the union of all visited atoms,
and the latter corresponds to the product of $p_i$ entries on the
visited subterms of probabilistic elements. Making these ideas
concrete, our example yields the following flat state distribution:

\small
\begin{tabular}{cccc}
\hline
Probability & (is-in R2D2)  & (is-in box)  & (is-in cup) \\
\hline
.24 & kitchen & office & office \\
.06 & kitchen & kitchen & office \\
.56 & kitchen & office & kitchen \\
.14 & kitchen & kitchen & kitchen \\
\hline
\end{tabular}
\normalsize

\section{Switching Planner}

We now describe a continual planning approach that switches between
sequential and decision-theoretic planning.

In this work, the problem at hand is a POMDP described in DTPDDL. From
hereon, we restrict our attention to POMDPs with deterministic
actions. Therefore it is worthwhile noting that any POMDP with
stochastic actions can be compiled into an equivalent
deterministic-action POMDP, where all the original action uncertainty
is expressed in the starting-state
distribution~\cite{ng:Jordan:2000}. In our setting, that of
finite-horizon planning, such a compilation yields a finite flat
representation of the original POMDP, only with deterministic
actions.

\Omit{
  Therefore, although our current implementation does not
  explicitly support it, our approach generalises to POMDPs with
  stochastic action models.
}

\subsection{Sequential Planning in POMDPs}

Our approach uses sequential planning to compute a linear plan, and
one trace of that plan's execution. The latter encapsulates
assumptions about: (1) the true underlying state, (2) how the plan's
execution will progress, and (3) the possibility of eventually holding
strong beliefs about the truth values of specific state
propositions. Sequential planning proceeds in a continual manner until
the verification of an assumption about a runtime state proposition is
scheduled. At that point the system switches to a decision-theoretic
planner that is used to tailor sensory processing by the robot to test
the validity of the assumptions made so far.

In a deterministic-action POMDP, all the uncertainty in state and
action is expressed in the $(\pp{:init})$ declaration. Our approach
uses the structure of that, as it occurs in the problem description,
do define a set of abstract state-assumptions available to 
sequential planning.  
%%
Writing \#\ if the value of a proposition is unspecified in the
assumption, for the $\pp{R2D2}$ example we have that the following
assumptions are available:

\small
\begin{tabular}{cccc}
\hline
Probability & (is-in R2D2)  & (is-in box)  & (is-in cup) \\
\hline
.24 & kitchen & office & office \\
.3 & kitchen & \# & office \\
.8 & kitchen & office & \# \\
.2 & kitchen & kitchen & \# \\
.7 & kitchen & \# &  kitchen\\
.06 & kitchen & kitchen & office \\
.56 & kitchen & office & kitchen \\
.14 & kitchen & kitchen & kitchen \\
1.0 & kitchen & \# & \# \\
\hline
\end{tabular}
\normalsize

\noindent A valid assumption in our setting corresponds to a {\em
relaxed} visitation of the root term of $(:init)$. In this relaxed
case, a conjunctive term is visited iff its atomic subterms are
visited, and zero or more of its immediate probabilistic subterms are
visited. The starting state, $s_0$, for sequential planning is the
abstract state with probability $1.0$ -- E.G., for $\pp{R2D2}$ $s_0
\equiv \{(= (is-in R2D2) kitchen)\}$.

In order to make assumptions available to the sequential planner, we
add one action $\actions(p_i;(T_i))$ to the problem for each element,
$p_i (T_i)$, of each probabilistic term from $(:init)$. The physics
for these actions is as follows. $\actions(p_i;(T_i))$ can be executed
if no $\actions(p_j;(T_j))$, $j \neq i$, has been executed from the
same probabilistic term, and, either
$(\pp{probabilistic}~..p_i~(T_i)..)$ is in the root conjunct, or it
occurs in $T_k$ for some executed $\actions(p_k;(T_k))$.
%%
Executing $\actions(p_i;(T_i))$ in $s$ effects a transition to a
successor state $s^v$ with probability $p_i$, and $s^\bot$ with
probability $1 - p_i$. Here, $s^v$ is characterised by the union of
$s$ and atomic terms from $T_i$, and $s^\bot$ is a sink state -- I.e.,
$\forall\action\in\actions$ $\reward(\state^\bot, \action) = 0$ and
$\transProb(\state^\bot, \action, \state^\bot) = 1$.

\subsection{Decision-Theoretic Planning in Abstractions}

The problem posed to the decision-theoretic system is an abstraction
of the problem at hand with a distinct utility model that focuses the
agents activities on assumption verification. The abstraction is
induced according to the executing sequential plan. Specifically, we
first omit all problem variables that are not mentioned in the linear
trace, and then refine that preliminary model, by progressively adding
more problem variables while the resulting state space remains of a
practicable size. Refinement in our approach is a simple and
inexpensive step, guided by the level at which variable symbols occur
in the description of the initial conditions. The abstract process is
also given a utility model that captures the reward (resp. penalty) of
correctly (resp. falsely) in/validating the assumptions made in the
linear trace. In detail, for each assumption on a state variable, we
can have a reward of $1$ for correctly rejecting it, and a penalty of
$-1$ for falsely rejecting it. Decision theoretic planning proceeds in
a more-or-less continual manner until an assumption is rejected, or
until it returns control to the sequential system because it can see
no benefit from performing further actions in the abstraction.

Addressing briefly the specifics of our mobile robot's
architecture. In our scenario the utility model for the POMDP posed to
decision-theoretic planning is sourced from a motivational component
of our robotic architecture. This proposes a non-zero reward
(resp. penalty) for sequential planning with a true (resp. false)
assumption about a state proposition.

@@

\section{Discussion and Related Work}

There have been a number of developments recently towards planning
under uncertainty using systems that were intended for sequential
planning in deterministic problems.  Notably,
\system{FFR$_a$}~\cite{yoon:etal:2007}, the winning entry from the
probabilistic track of the 2004 International Planning Competition.
In the continual paradigm, \system{FFR$_a$} uses the fast satisficing
procedure \system{FF}~\cite{hoffmann:nebel:2001} to compute sequential
plans and corresponding execution traces.
%%
More computationally expensive approaches in this vein combine
sampling strategies on valuations over {\em runtime variables} with
deterministic planning procedures. The outcome is typically a more
robust sequential plan~\cite{yoon:etal:2008}, or contingent
plan~\cite{majercik:2006}.
%%
Also leveraging deterministic planners in problems that feature
uncertainty, \system{Conformant-FF}~\cite{hoffmann:brafman:2006} and
$T_0$~\cite{palacios:geffner:2009} demonstrate how conformant planning
---i.e., sequential planning in unobservable worlds--- can be modelled
as a deterministic problem, and therefore solved using sequential
systems. In this conformant setting, advances have been towards
compact representations of beliefs amenable to existing best-first
search planning procedures, and lazy evaluations of beliefs. Most
recently this research thread has been extended to contingent planning
in fully observable non-deterministic
environments~\cite{albore:etal:2009}.

@@

@@

Continual planning is a powerful technique that goes some way to
addressing those challenges. That approach interleaves planning and
execution, deliberately postponing planning for contingencies unless
they eventuate during execution.

computing a single sequential plan and
eventuality

@@

%% More generally, a useful
%% formalism for representing solutions to POMDPs is the finite-state
%% controller (FSC). This is a three-tuple $\langle \nodes, \selection,
%% \transition \rangle$ where: $\node \in \nodes$ is a set of nodes,
%% $\selection_\node(\action) = P(\action | \node)$, and
%% $\transition_\node(\action, \observation, \node') = P(\node'|\node,
%% \action, \observation)$. Where the value of acting is the discounted
%% accumulated reward over an infinite horizon, then each controller can
%% be assigned a value, written $V_\node(\state)$, expressed in the usual
%% way according to the Bellman equation:

%% \begin{equation}\label{eq:evaluation}
%% \begin{array}{lcl}
%% V_\node(\state) & = & \sum_{\action \in \actions}
%% \selection_\node(\action) \reward(\state, \action) \;\; + \vspace{1ex} \\

%% && \hspace{-10ex} \beta \sum_{\action, \observation,
%% \state', \node'} \transition_\node(\action, \observation, \node')
%% \transProb(\state, \action, \state') \obsDist_\observ(\state',
%% \action) V_{\node'}(\state')
%% \end{array}
%% \end{equation}

%% \noindent Here, the value of {\em prior} $\bstate$ given an FSC is
%% then:

%% \begin{equation} \label{eq:valueBelief}
%% V_{\pp{FSC}}(\bstate) = \max_{\node \in \nodes} \sum_{\state \in \states} \bstate(\state) V_\node(\state)
%% \end{equation}

%% \noindent There is a corresponding deterministic FSCs for both
%% sequential (i.e., conformant) and contingent
%% plans. Eq~\ref{eq:evaluation} gives their value in the finite-horizon
%% case, if we take $\beta = 1$, and alter the process dynamics so there
%% is a compulsory transition to a zero utility sink state when that
%% horizon is reached.

Sequential planning in our approach proceeds in two phases. In the
first phase, the goal is to identify a plan prefix

In sequential planning, the plan prefix corresponds to choosing a
valid abstract starting state.

moreover, for the purpose of decision-theoretic abstractions,

A relaxed form of {\em visitation} we call {\em abstract
  visitation}. In the abstract case a conjunctive term is visited
  provided all its atomic subterms are visited, and zero or more of
  its probabilistic terms are visited.

The planner has a compulsory starting action, that visits the root
term, and then has to execute visitation actions until that plan
prefix corresponds to a valid {\em abstract visitation} of the {\em
:init} block. Beyond such a prefix, the process actions and senses are
available to the planner

Goal expressions in the problem description ---for example
$(\pp{kval}~(\pp{location}~\pp{box}))$ expresses ``the robot knows the
location of the {\em box}--- are treated as control-knowledge by the
planner. Here, a plan is not considered valid unless in the finial
stet the goal condition is satisfied.
